---
title: "Learning new-tidyr and a cheeky steal from purrr"
author: ''
date: '2020-03-02'
slug: pivot-function
categories:
- R
tags:
- R
- tidyr
- pivot
- purrr
summary: "A quick bit of new-tidyr / pivot learning, and a cheaky steal from purrr for good measure"
image:
  caption: ''
  focal_point: ''
output:
  blogdown::html_page:
    toc: false
    number_sections: false
    toc_depth: 1
draft: false
math: false
---

  
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.width=10, fig.height=6, fig.align = "center", 
                      dev.args = list(png = list(type = "cairo")), fig.retina=3) 

library(Cairo)

set.seed(123)

```


# Ah, pivots...

So, 'pivots' have been part of my life in data analysis since my first interview, where I missed the point of the test (to make a pivot) and manually filtered loads of Excel data until I got there.  Managed to get the job though (nothing to do with the sympathy vote for the broken shoulder I had at the time, of course ...).

They are so easy to build in Excel, a real pain in SQL (as you need to specify the column names, or make it dynamic), but thank heavens for the `tidyr` package.  Now, I'm not the first to suggest it could be a bit confusing, but I got my head around `spread()` and `gather()` and was a happy bunny.  Then they changed it... Hadley Wickham and contributors made the wise choice to adopt the phrase 'pivot' in `pivot_wider()` and `pivot_longer()`.  So I felt like a right idiot when I couldn't figure it out.... I took the sensible approach of ignoring it until they forced my hand by removing it.

That said, I was working on a solution in some consultancy work today, and I needed to pivot the data but also to leave a strong markdown documentation trail.  I felt it would have been a bit unfair if I left them with `spread()` knowing it would go the way of the Dodo soon.  This brief post is just showing a working bit of pivot for my own memory and anyone else who's interested.

I'm pivoting the `NHSRdatasets` `LOS_models` data.frame for organisation by columns, and applying function.  This could be `mean()` or `meadian()` or similar, but I defined a function to add some Poisson distributed noise.

```{r pivot1, message=FALSE, warning=FALSE}

library(NHSRdatasets)

data("LOS_model")

library(tidyr)
library(dplyr)

set.seed(123) # For reproducibility

LOS_model %>% 
  pivot_wider(id_cols = ID
              , names_from = Organisation
              , values_from = LOS
              , values_fill = list(LOS = 0)
              , values_fn = list(LOS= function(x){x+ rpois(n = 1, lambda = 3)})
              )


```

## Bonus round:

I remember hearing Tom Jemmet [@tomjemmet](https://twitter.com/tomjemmett) talk about the magic of `purrr`.  Now I'm a bit slow off the mark with `purrr` because I couldn't see why it was different to `apply` functions.  Now I've hear more about it, I can see there's more consistency and it allows some nice applications of functions.  The thing that was _really_ nice though was ___anonymous functions___! They look so much cleaner, and are defined as a one-sided formula.  I've replaced the aggregate function with an anonymous one in the example below:

```{r pivot2, message=FALSE, warning=FALSE}
library(purrr)

set.seed(123) # For reproducibility

LOS_model %>% 
  pivot_wider(id_cols = ID
              , names_from = Organisation
              , values_from = LOS
              , values_fill = list(LOS = 0)
              , values_fn = list(LOS= ~ . + rpois(n = 1, lambda = 3))
  )

```


Maybe that's of use to you.  Hope so!
